%*******************************************************
% Acknowledgments
%*******************************************************
\pdfbookmark[1]{Acknowledgements}{acknowledgements}

\bigskip

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax
\chapter*{Acknowledgements}

\def\thanks#1{%
\begingroup
\leftskip1em
\noindent #1
\par
\endgroup
}

This thesis would not have been possible to complete without the support and guidance that I received from many wonderful individuals to whom I want to say \emph{thank you}.

\vspace{0.12cm}
\noindent I am deeply grateful to my supervisor from the Wrocław University of Science and Technology, Arkadiusz Wójs, for encouraging me to work in the field of condensed matter physics and accepting me as a PhD student. His good advices and assistance, especially at the beginning of my scientific career, have been invaluable.

\vspace{0.12cm}
\noindent I am fully indebted to Titus Mangham-Neupert from the University of Zurich for giving me the life-changing opportunity to pursue the research in his group. His enthusiasm, creativity and wisdom are inspirational. I do appreciate his endless patience and continued support of my work.

\vspace{0.12cm}
\noindent I would like to acknowledge Paweł Potasz for helping me during my master studies and first years of doctorate. His experience in numerical methods and willingness to share the knowledge were crucial at early stages of my serious coding.

\vspace{0.12cm}
\noindent My appreciation extends to my colleagues from UZH. I have benefited a lot from our joint projects and not work-related discussions. In particular, I would like to thank Frank for having fruitful collaborations and great time during many conferences.

\vspace{0.12cm}
\noindent Living in Switzerland would have been more complicated without the assistance of the administrative staff at UZH. I am especially grateful to Gaby Aeppli-Koller and Brigitte Freund for helping me with all problems related to the Swiss bureaucracy. I also cannot forget about the people outside the workplace that helped me in settling down in Zurich: Samuel, Antonio and Martin.

\vspace{0.12cm}
\noindent Many thanks to my fellow doctoral students from Wrocław, with a special mention of Maciek and Adam; the laughs we shared will be remembered. I would like to express my heartfelt gratitude to Karolina for our friendship and being someone I can count on.

\vspace{0.12cm}
\noindent Thank you, Beata, for being a close friend, despite the kilometres that separate us. I hope we will see each other more often soon!

\vspace{0.12cm}
\noindent I cannot thank Kasia enough for our long-lasting friendship. Thanks for being supportive and helping me in surviving the ups and downs all these years.

\vspace{0.12cm}
\noindent I owe a great debt of gratitude to Lo\"ic, who carefully proofread my thesis and gave a lot of helpful suggestions. I thank you so much for the incredible moral support over the last months.

\vspace{0.12cm}
\noindent Above all, I would like to thank my family for their endless love and not letting me give up. 

\endgroup
